---
title: "Store Deliveries"
author: "Anil Kumar"
date: "7/31/2021"
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1) Introduction

The business problem is that at the beginning of the COVID lockdown a local company that faciliates ordering and home delivery of products has become inundated with orders.  As a result of this increased demand, many orders are being voided by participating stores and there are also many deliveries that take longer than the 1 hour goal.  The data is from a one week period at the start of the COVID lockdown.  The company is trying to gain more insight about the factors that contribute to orders getting voided and taking longer to deliver.  The company utilizes three different stores that supply and delivery the products.    

## 2) Pre-processing and Cleaning

```{r read in the necessary libraries, message=FALSE}
# Include the necessary libraries 
# (not all may be used...but I usually include the most common ones I use by habit!)
library(tidyverse)
library(readxl)
library(jtools)
library(plotly)
library(caret)
library(pROC)
library(MLmetrics)
library(janitor)
library(lubridate)
```


```{r read in csv file and clean column names}
#read in the data and clean the column names
df_raw= read_excel("Store_Deliveries2.xlsx", sheet = 1, na='N/A') 
df_raw <- df_raw %>% clean_names()
```


```{r examine column names and types}
#examine dataset
glimpse(df_raw)
```

```{r summary of df columns and count N/A}
#examine individual columns and count the number of missing values
summary(df_raw)
sum(is.na(df_raw))
```

A few findings from inspecting the summary:

* There are 318 missing values and from the summary these are all from the 'delivery_time' column
* There is only one value (4.99) for the 'delivery_fee' column
* There is only one value (1) for the 'delivery_neighborhood_id' column

Next, the dataset is manipulated and cleaned prior to analysis and visualization.
```{r clean dataframe}
#clean the raw data and appropriately rename columns 

data <- df_raw %>% 
  
  #Drop non-predictive variables
  select(-delivery_neighborhood_id, -order_id, -delivery_fee) %>% 
  
  # make retailer_id into factor and rename column
  mutate(retailer_id = factor(retailer_id)) %>% 
  mutate(retailer_id = factor(retailer_id, labels=c("Store A","Store B","Store C"))) %>%
  rename(Store = retailer_id) %>%
  
  # find day of week and label appropriately 
  mutate(Day = factor(weekdays(order_date))) %>% 
  mutate(Day = factor(Day, levels=c("Monday", "Tuesday","Wednesday",
                                    "Thursday","Friday", "Saturday","Sunday"))) %>% 
  select(-order_date) %>% # drop original column as no longer need
                 
  # turn substitution column into factor and rename
  mutate(substitution = factor(ifelse(substitution==TRUE,
                               "1 or more substitution(s)","No subsitution(s)"))) %>% 
  rename(Substitution = substitution) %>%
                
  # rename GMV column
  rename(GMV = gmv) %>% 
  
  #rename delivery_time column
  rename(Delivery_Time = delivery_time) %>% 
  
  # make order_status into factor 
  mutate(order_status = factor(order_status)) %>% 
  rename(Order_Status = order_status) %>% 
                 
  # Parse hour of order_time and label appropriately 
  mutate(Hour = factor(substring(as.character(order_time),12,13))) %>% 
  mutate(Hour = factor(Hour, labels=c("12pm", "1pm","2pm","3pm","4pm", 
                                      "5pm","6pm","7pm","8pm"))) %>%
  select(-order_time) %>% # drop original column as no longer needed
  
  # Binarize Hour variable 
  mutate(Time_of_Day = ifelse(Hour %in% c('12pm','1pm','2pm','3pm','4pm'),
                                         "Earlier than 5pm","Later than 5pm")) %>%
  
  # create column for past 1hr delivery
  mutate(Delivery_Over_1hr = factor(ifelse(Delivery_Time > 60,"Over 1hr", "1hr or less"))) %>% 
  
  # create column for weekday vs. weekend
  mutate(Weekend = factor(ifelse(Day %in% c('Saturday','Sunday','Friday'),
                                           "Weekend","Non-Weekend"))) 
```

The data is re-examined to make sure cleaning was performed appropriately.  
```{r re-examine data to make sure columns are correct types}
summary(data)
```

From the above summaries, it is apparent that the missing values are for deliveries that were voided (318 voided orders in total).  The 318 missing values for the 'Delivery_Over_1hr' and 'Delivery_Time' columns are due to the fact that these are observations are voided orders, and therefore don't have an associated delivery time. 

**DATA DICTIONARY:**

* **Store:** Represents the store; Each store is open daily from 12pm-9pm
* **Order_Status:** _Delivered_ = Order was delivered successfully; _Void_ = Order was voided (cancelled) by retailer; The data does not include orders that were voided by the consumer as the number was negligible
* **Delivery_Time:** The number of minutes between when the order was placed and when the order was delivered
* **GMV	(Gross Merchandise Value):** the dollar value of products that was ordered
* **Substitution:** Whether he retailer substituted at least one item on a consumer's order due to availability
* **Day:** Day of week order was placed
* **Weekend:** Whether order was placed on a weekend or non-weekend
* **Hour:** Hour order was placed (12pm = 12:00pm-12:59pm, 1pm = 1:00pm-1:59pm, etc.)
* **Time_of_Day:** Whether order was placed before 5pm or after 5pm
* **Delivery_Over_1hr:** Whether order was delivered in over 1hour vs. 1hr or less

## 3) Exploratory Data Analysis 

Below, features are visually explored to assess their associations with orders being voided and longer delivery times.  I have also created an interactive dashboard using Tableau to assess these relationships.  Feel free to use and play around with the dashboard here:
https://public.tableau.com/app/profile/anil.kumar8305/viz/OrderDeliveryTimesandVoids/VoidorLate

### 3a) Exploring Order Status

First, voided orders are explored in order to examine any trends in the data.  Of all of the 2320 orders, 318 were voided by a retailer (13.7%)
```{r Order Status counts}
data %>% group_by(Order_Status) %>% summarize(count=n()) %>%
         mutate(freq = round(count / sum(count),3)) 
```

Now, the order status is explored further with resepect to the other variables.  

```{r Order Status and Store, message=FALSE}
data %>% group_by(Store, Order_Status) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count / sum(count),3)) %>%
         ggplot(aes(x=Store, y=Percent, fill=Order_Status)) + geom_col(position = "dodge") + 
         scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) +
         ggtitle("Percentage of Voided Orders by Store") +
         theme_bw()
```


```{r Order Status and Day/Weekend, message=FALSE}
data %>% group_by(Day, Order_Status) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count / sum(count),3)) %>%
         ggplot(aes(x=Day, y=Percent, fill=Order_Status)) + geom_col(position = "dodge") + 
         scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) +
         ggtitle("Percentage of Voided Orders by Day") +
         theme_bw()

data %>% group_by(Weekend, Order_Status) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count / sum(count),3)) %>%
         ggplot(aes(x=Weekend, y=Percent, fill=Order_Status)) + geom_col(position = "dodge") + 
         scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) +
         ggtitle("Percentage of Voided Orders by Weekend") +
         theme_bw()
```


```{r Order Status and Hour/Time_of_Day, message=FALSE}
data %>% group_by(Hour, Order_Status) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count / sum(count),3)) %>%
         ggplot(aes(x=Hour, y=Percent, fill=Order_Status)) + geom_col(position = "dodge") + 
         scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) +
         ggtitle("Percentage of Voided Orders by Hour") +
         theme_bw()

data %>% group_by(Time_of_Day, Order_Status) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count / sum(count),3)) %>%
         ggplot(aes(x=Time_of_Day, y=Percent, fill=Order_Status)) + geom_col(position = "dodge") + 
         scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) +
         ggtitle("Percentage of Voided Orders by Time of Day") +
         theme_bw()
```


```{r Order Status and Substitution, message=FALSE}
data %>% group_by(Substitution, Order_Status) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count / sum(count),3)) %>%
         ggplot(aes(x=Substitution, y=Percent, fill=Order_Status)) + geom_col(position = "dodge") + 
         scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) +
         ggtitle("Percentage of Voided Orders by Substitution") +
         theme_bw()
```

```{r Order Status and GMV}
temp <- data %>%  group_by(Order_Status) %>% 
        summarize(mean_GMV = round(mean(GMV, na.rm=TRUE),2), 
                  median_GMV=round(median(GMV, na.rm=TRUE),2))

data %>% ggplot(.,aes(x=GMV)) + geom_density() + facet_grid(rows=vars(Order_Status)) + 
          theme_bw() + 
          labs(x="GMV") +
          scale_x_continuous(labels = function(x) paste0("$",x)) +
          geom_vline(aes(xintercept=20), color='black', linetype='dashed') +
          geom_vline(data=temp, aes(xintercept=mean_GMV), color='red') +
          geom_vline(data=temp, aes(xintercept=median_GMV), color='blue') + 
          ggtitle("Distributions of GMV on Voided vs. Delivered Orders", 
                  subtitle="Dashed Black Reference Line = $20") +
          geom_text(data=temp,aes(x=190, label=paste("Median:",as.character(median_GMV)), 
                                  y=0.019), colour="blue",vjust = 1.5)+
          geom_text(data=temp,aes(x=190, label=paste("Mean:",as.character(mean_GMV)),
                                  y=0.03), colour="red",vjust = 1.5)

data %>% ggplot(.,aes(x=Order_Status, y=GMV, fill=Order_Status)) + geom_boxplot() +
          scale_y_continuous(labels = function(y) paste0("$",y)) +
          geom_hline(aes(yintercept=20), color='black', linetype='dashed') +
          ggtitle("Distributions of GMV on Voided vs. Delivered Orders", 
                  subtitle="Dashed Black Reference Line = $20") +
          theme_bw()
```

Ther are some interesting trends from the above visualizations.  Voided orders occurred:

* ONLY for orders placed after 5pm
* ONLY on orders where items subsitutions were not permitted
* Mostly on weekends
* Mostly on orders with lower GMV's

Let's visually explore the interaction between the Weekend/Time_of_Day and the relationship to Order_Status.  From the figure below, it'c clear that voided orders ONLY occur after 5pm, and the vast majority are on weekends.
```{r Interaction between Weekend/Hour and the relationship to Order Status, message=FALSE}
 
data %>% group_by(Weekend, Time_of_Day, Order_Status) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count/sum(count),2)) %>%
         ggplot(., aes(x=Time_of_Day,y=Percent, fill=Order_Status)) + geom_col(position = "dodge") + 
               scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) +
               facet_grid(rows= vars(Weekend)) +
               ggtitle("Percentage of Voided Orders by Weekend and Time of Day") +
               theme_bw()
```

### 3b) Exploring Delivery Time

Of all of the 2002 orders that were delivered, 1,037 (51.8%) took over 1 hour to deliver.  The median delivery time is 63 minutes and the mean is 77.05 minutes.  
```{r Delivery_Over_1hr counts}
data %>% drop_na() %>%
         group_by(Delivery_Over_1hr) %>% summarize(count=n()) %>%
         mutate(freq = round(count / sum(count),3)) 

data %>% drop_na() %>%
         summarize(mean_delivery_time=mean(Delivery_Time), 
                   median_delivery_time=median(Delivery_Time) )
```


```{r Delivery Time and Retailer, message=FALSE}
data %>% drop_na() %>%
         group_by(Store, Delivery_Over_1hr) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count / sum(count),2)) %>%
         ggplot(., aes(x=Store,y=Percent, fill=Delivery_Over_1hr)) + geom_col(position = "dodge") +
         scale_y_continuous(labels = function(x) paste0(x*100, "%"),limits = c(0,1)) + 
         ggtitle("Percentage of Delivery Times Over 1hr by Store") +
         theme_bw()

data %>%  drop_na() %>%
          ggplot(.,aes(x=Store, y=Delivery_Time, fill=Store)) + 
          geom_boxplot() +
          geom_hline(aes(yintercept=60), color='black', linetype='dashed') + 
          ggtitle("Distributions of Delivery Times by Store", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          theme_bw() 

temp <- data %>%  group_by(Store) %>% 
        summarize(mean_delivery_time = round(mean(Delivery_Time, na.rm=TRUE),2), 
                 median_delivery_time=round(median(Delivery_Time, na.rm=TRUE),2))

data %>% drop_na() %>%
          ggplot(.,aes(x=Delivery_Time)) + geom_density() + 
          facet_wrap(~Store) + 
          theme_bw() + 
          labs(x="Delivery Time (Minutes)") +
          geom_vline(aes(xintercept=60), color='black', linetype='dashed') +
          geom_vline(data=temp, aes(xintercept=mean_delivery_time), color='red') +
          geom_vline(data=temp, aes(xintercept=median_delivery_time), color='blue') + 
          ggtitle("Distributions of Delivery Times by Store", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          geom_text(data=temp,aes(x=190, label=paste("Median:",as.character(median_delivery_time)), 
                                  y=0.006), colour="blue",vjust = 1.5)+
          geom_text(data=temp,aes(x=190, label=paste("Mean:",as.character(mean_delivery_time)),
                                  y=0.007), colour="red",vjust = 1.5)

```


```{r Delivery Time and Day/Weekend, message=FALSE}
data %>% drop_na() %>%
         group_by(Day, Delivery_Over_1hr) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count / sum(count),2)) %>%
         ggplot(., aes(x=Day,y=Percent, fill=Delivery_Over_1hr)) + geom_col(position = "dodge") +
         scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) + 
         ggtitle("Percentage of Delivery Times Over 1hr by Day") + 
         theme_bw()

data %>% drop_na() %>%
         group_by(Weekend, Delivery_Over_1hr) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count / sum(count),2)) %>%
         ggplot(., aes(x=Weekend,y=Percent, fill=Delivery_Over_1hr)) + geom_col(position = "dodge") +
         scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) + 
         ggtitle("Percentage of Delivery Times Over 1hr by Weekend") + 
         theme_bw()

data %>%  drop_na() %>%
          ggplot(.,aes(x=Day, y=Delivery_Time, fill=Day)) + 
          geom_boxplot() +
          geom_hline(aes(yintercept=60), color='black', linetype='dashed') + 
          ggtitle("Distributions of Delivery Times by Day", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          theme_bw()  

data %>%  drop_na() %>%
          ggplot(.,aes(x=Weekend, y=Delivery_Time, fill=Weekend)) + 
          geom_boxplot() +
          geom_hline(aes(yintercept=60), color='black', linetype='dashed') + 
          ggtitle("Distributions of Delivery Times by Weekend", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          theme_bw() 

temp <- data %>% group_by(Day) %>% 
        summarize(mean_delivery_time = round(mean(Delivery_Time, na.rm=TRUE),2), 
                  median_delivery_time=round(median(Delivery_Time, na.rm=TRUE),2))
     

data %>% drop_na() %>%
          ggplot(.,aes(x=Delivery_Time)) + geom_density() + 
          facet_wrap(~Day) + 
          theme_bw() + 
          labs(x="Delivery Time (Minutes)") +
          geom_vline(aes(xintercept=60), color='black', linetype='dashed') +
          geom_vline(data=temp, aes(xintercept=mean_delivery_time), color='red') +
          geom_vline(data=temp, aes(xintercept=median_delivery_time), color='blue') + 
          ggtitle("Distributions of Delivery Times by Day", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          geom_text(data=temp,aes(x=190, label=paste("Median:",as.character(median_delivery_time)), 
                                  y=0.019), colour="blue",vjust = 1.5)+
          geom_text(data=temp,aes(x=190, label=paste("Mean:",as.character(mean_delivery_time)),
                                  y=0.03), colour="red",vjust = 1.5)


temp <- data %>%  group_by(Weekend) %>% 
        summarize(mean_delivery_time = round(mean(Delivery_Time, na.rm=TRUE),2), 
                  median_delivery_time=round(median(Delivery_Time, na.rm=TRUE),2))
     

data %>% drop_na() %>%
          ggplot(.,aes(x=Delivery_Time)) + geom_density() + 
          facet_wrap(~Weekend) + 
          theme_bw() + 
          labs(x="Delivery Time (Minutes)") +
          geom_vline(aes(xintercept=60), color='black', linetype='dashed') +
          geom_vline(data=temp, aes(xintercept=mean_delivery_time), color='red') +
          geom_vline(data=temp, aes(xintercept=median_delivery_time), color='blue') + 
          ggtitle("Distributions of Delivery Times by Weekend", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          geom_text(data=temp,aes(x=190, label=paste("Median:",as.character(median_delivery_time)), 
                                  y=0.019), colour="blue",vjust = 1.5)+
          geom_text(data=temp,aes(x=190, label=paste("Mean:",as.character(mean_delivery_time)),
                                  y=0.022), colour="red",vjust = 1.5)
```


```{r Delivery Time and Hour/Time_of_Day, message=FALSE}
data %>% drop_na() %>%
          group_by(Hour, Delivery_Over_1hr) %>% 
          summarize(count=n()) %>%
          mutate(Percent = round(count / sum(count),2)) %>%
          ggplot(., aes(x=Hour,y=Percent, fill=Delivery_Over_1hr)) + geom_col(position = "dodge") +
          scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) + 
          ggtitle("Percentage of Delivery Times Over 1hr by Hour") + 
          theme_bw()

data %>% drop_na() %>%
          group_by(Time_of_Day, Delivery_Over_1hr) %>% 
          summarize(count=n()) %>%
          mutate(Percent = round(count / sum(count),2)) %>%
          ggplot(., aes(x=Time_of_Day,y=Percent, fill=Delivery_Over_1hr)) + geom_col(position = "dodge") +
          scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) + 
          ggtitle("Percentage of Delivery Times Over 1hr by Time of Day") + 
          theme_bw()

data %>%  drop_na() %>%
          ggplot(.,aes(x=Hour, y=Delivery_Time, fill=Hour)) + 
          geom_boxplot() +
          geom_hline(aes(yintercept=60), color='black', linetype='dashed') + 
          ggtitle("Distributions of Delivery Times by Hour", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          theme_bw() 

data %>%  drop_na() %>%
          ggplot(.,aes(x=Time_of_Day, y=Delivery_Time, fill=Time_of_Day)) + 
          geom_boxplot() +
          geom_hline(aes(yintercept=60), color='black', linetype='dashed') + 
          ggtitle("Distributions of Delivery Times by Time of Day", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          theme_bw() 

temp <- data %>% group_by(Hour) %>% 
        summarize(mean_delivery_time = round(mean(Delivery_Time, na.rm=TRUE),2), 
                  median_delivery_time=round(median(Delivery_Time, na.rm=TRUE),2))

data %>% drop_na() %>%
          ggplot(.,aes(x=Delivery_Time)) + geom_density() + 
          facet_wrap(~Hour) + 
          theme_bw() + 
          labs(x="Delivery Time (Minutes)") +
          geom_vline(aes(xintercept=60), color='black', linetype='dashed') +
          geom_vline(data=temp, aes(xintercept=mean_delivery_time), color='red') +
          geom_vline(data=temp, aes(xintercept=median_delivery_time), color='blue') + 
          ggtitle("Distributions of Delivery Times by Hour", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          geom_text(data=temp, aes(x=190, label=paste("Median:",as.character(median_delivery_time)), 
                                   y=0.027), colour="blue",vjust = 1.5)+
          geom_text(data=temp, aes(x=190, label=paste("Mean:",as.character(mean_delivery_time)), 
                                   y=0.04), colour="red",vjust = 1.5)

temp <- data %>% group_by(Time_of_Day) %>% 
        summarize(mean_delivery_time = round(mean(Delivery_Time, na.rm=TRUE),2), 
                  median_delivery_time=round(median(Delivery_Time, na.rm=TRUE),2))

data %>% drop_na() %>%
          ggplot(.,aes(x=Delivery_Time)) + geom_density() + 
          facet_wrap(~Time_of_Day) + 
          theme_bw() + 
          labs(x="Delivery Time (Minutes)") +
          geom_vline(aes(xintercept=60), color='black', linetype='dashed') +
          geom_vline(data=temp, aes(xintercept=mean_delivery_time), color='red') +
          geom_vline(data=temp, aes(xintercept=median_delivery_time), color='blue') + 
          ggtitle("Distributions of Delivery Times by Time of Day", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          geom_text(data=temp, aes(x=190, label=paste("Median:",as.character(median_delivery_time)), 
                                   y=0.035), colour="blue",vjust = 1.5)+
          geom_text(data=temp, aes(x=190, label=paste("Mean:",as.character(mean_delivery_time)), 
                                   y=0.04), colour="red",vjust = 1.5)
```


```{r Delivery Time and Substitution, message=FALSE}
data %>% drop_na() %>%
          group_by(Substitution, Delivery_Over_1hr) %>% 
          summarize(count=n()) %>%
          mutate(Percent = round(count / sum(count),2)) %>%
          ggplot(., aes(x=Substitution,y=Percent, fill=Delivery_Over_1hr)) + geom_col(position = "dodge") +
          scale_y_continuous(labels = function(x) paste0(x*100, "%"), limits = c(0,1)) + 
          ggtitle("Percentage of Delivery Times Over 1hr by Substitution") +
          theme_bw()

data %>%  drop_na() %>%
          ggplot(.,aes(x=Substitution, y=Delivery_Time, fill=Substitution)) + 
          geom_boxplot() +
          geom_hline(aes(yintercept=60), color='black', linetype='dashed') + 
          ggtitle("Distributions of Delivery Times by Substitution", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          theme_bw()

temp <- data %>% group_by(Substitution) %>% 
        summarize(mean_delivery_time = round(mean(Delivery_Time, na.rm=TRUE),2), 
                  median_delivery_time=round(median(Delivery_Time, na.rm=TRUE),2))

data %>% drop_na() %>%
          ggplot(.,aes(x=Delivery_Time)) + geom_density() + 
          facet_wrap(~Substitution) + 
          theme_bw() + 
          labs(x="Delivery Time (Minutes)") +
          geom_vline(aes(xintercept=60), color='black', linetype='dashed') +
          geom_vline(data=temp, aes(xintercept=mean_delivery_time), color='red') +
          geom_vline(data=temp, aes(xintercept=median_delivery_time), color='blue') + 
          ggtitle("Distributions of Delivery Times by Substitution", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          geom_text(data=temp, aes(x=190, label=paste("Median:",as.character(median_delivery_time)), 
                                   y=0.009), colour="blue",vjust = 1.5)+
          geom_text(data=temp, aes(x=190, label=paste("Mean:",as.character(mean_delivery_time)), 
                                   y=0.010), colour="red",vjust = 1.5)
```


```{r Delivery Time and GMV, message=FALSE}
data %>% drop_na() %>%
        ggplot(., aes(x=GMV,y=Delivery_Time, shape=Time_of_Day, color=Weekend)) + 
               geom_jitter(alpha=0.7) +
               geom_hline(aes(yintercept=60), color='black', linetype='dashed') + 
               ggtitle("Scatterplot of of Delivery Times vs. GMV", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
               theme_bw()
```

Similar to voided orders, the above visualizations shows that orders that delivery times are higher after 5pm and on weekends.  Also, the stores seem to have varying distributions of delivery times, with Store A being the slowest and Store C the quickest.  

Let's visually explore the interaction between the Weekend/Time_of_Day and the relationship to Delivery_Time.  From the figures below, it'c clear that orders placed after 5pm on weekends have a high chance of having a delivery time over 1hr.  Regardless of the day, any orders placed before 5pm are delivered within an hour. 

```{r Intercation between Weekend/Hour and the relationship to Order Status, message=FALSE}
data %>% drop_na() %>%
         group_by(Weekend, Time_of_Day, Delivery_Over_1hr) %>% 
         summarize(count=n()) %>%
         mutate(Percent = round(count/sum(count),2)) %>%
         ggplot(., aes(x=Time_of_Day,y=Percent, fill=Delivery_Over_1hr)) + geom_col(position = "dodge") + 
               scale_y_continuous(labels = function(x) paste0(x*100, "%")) + 
               ggtitle("Percentage of Delivery Times Over 1hr by Weekend and Time of Day") +
               facet_grid(rows= vars(Weekend)) +
               theme_bw()

data %>%  drop_na() %>%
          ggplot(.,aes(x=Hour, y=Delivery_Time, fill=Weekend)) + 
          geom_boxplot() +
          geom_hline(aes(yintercept=60), color='black', linetype='dashed') + 
          ggtitle("Distributions of Delivery Times by by Weekend and Hour", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          theme_bw()

temp <- data %>% group_by(Weekend, Time_of_Day) %>% 
        summarize(mean_delivery_time = round(mean(Delivery_Time, na.rm=TRUE),2), 
                  median_delivery_time=round(median(Delivery_Time, na.rm=TRUE),2))

data %>% drop_na() %>%
          ggplot(.,aes(x=Delivery_Time)) + geom_density() + 
          facet_grid(rows=vars(Weekend), cols=vars(Time_of_Day)) + 
          theme_bw() + 
          labs(x="Delivery Time (Minutes)") +
          geom_vline(aes(xintercept=60), color='black', linetype='dashed') +
          geom_vline(data=temp, aes(xintercept=mean_delivery_time), color='red') +
          geom_vline(data=temp, aes(xintercept=median_delivery_time), color='blue') + 
          ggtitle("Distributions of Delivery Times by Weekend and Time of Day", 
                  subtitle="Dashed Black Reference Line = 60 minutes") +
          geom_text(data=temp, aes(x=190, label=paste("Median:",as.character(median_delivery_time)), 
                                   y=0.03), colour="blue",vjust = 1.5)+
          geom_text(data=temp, aes(x=190, label=paste("Mean:",as.character(mean_delivery_time)), 
                                   y=0.037), colour="red",vjust = 1.5)
```


From the exploratory data analysis, it's clear that orders placed on weekends and after 5pm are the biggest contributors to both orders being voided by retailers and delivery times being greater than 1 hour.  From, a business standpoint, this information is valuable as it shows where adjustments need to be made to avoid voids or longer delivery times.  For the purposes of this project, the majority of possible potential business solutions will not be discussed in detail.   

One solution that will be discussed is trying to accurately predict delivery times so that even if deliveries are going to take longer than one hour, at least an approximate time can be given to the customer.  This solution is discussed in the section below.    


## 4) Predicting Delivery Times

First, training and validation data sets are created.  The training set consists of 70% of the data and the validation set consists of 30% of the data.  A partition is created so that both the training and validation sets have an equal proportion of observations where the delivery time was over 1hr.

```{r split into training and validation sets}
set.seed(2)

temp <- data %>% drop_na() %>% 
                       select(Delivery_Time, Delivery_Over_1hr, Weekend, Store, Day, Hour,
                              Time_of_Day, GMV, Substitution)

index = createDataPartition(temp$Delivery_Over_1hr, p = 0.70, list = FALSE)
data_train = temp[index, ]
data_val = temp[-index, ]
```


Next, to check that the split was made appropriately.  This is especially important because the data was split on a binary variable and it the outcome being looked at is continuous.  Based on the overlying density plots and mean delivery times of both of the datasets, an appropriate split was made.  
```{r verify good split}
mean(data_train$Delivery_Time)
mean(data_val$Delivery_Time)

ggplot(data_train) + geom_density(data=data_train, aes(x=Delivery_Time, col='blue')) + 
                     geom_density(data=data_val, aes(x=Delivery_Time, col='red')) +
                     theme_bw() +
                     scale_color_manual(labels = c("Training", "Validation"), values = c("blue", "red")) +
                     guides(color=guide_legend("Dataset")) +
                     ggtitle("Distributions of of Delivery Times for Training and Validation datasets") 
```

Now, a linear regression model is created to predict Delivery_Time using Store, Day, Hour, GMV, and Substitution.  The model also includes a 3-way interaction term between Day/Hour/Store based on the EDA.
This model has an Adj. R^2^ of 0.897 compared to the Adj. R^2^ of 0.693 for the simpler model with no interactions terms (the output of this model is not shown).
```{r create linear model}
model <- lm(Delivery_Time ~ Store +
                            Substitution +
                            Day  +
                            Hour +
                            GMV +
                            Day:Hour:Store, 
                            data=data_train)

summ(model, digits=3, confint = TRUE)
```

Mean absolute errors for predicted delivery times using the model:
Training Set: 10.29 minutes
validation Set: 10.89 minutes
```{r predict times using training and validation sets and calculate Mean Absolute Error, warning=FALSE}
preds = predict(model, newdata=data_train)

data_train %>% select(Delivery_Time) %>%
               rename(Actual_Time=Delivery_Time) %>%
               mutate(Predicted_Time = preds) %>%
               mutate(abs_error=abs(Actual_Time-Predicted_Time)) %>%
               summarize(Mean_Absolute_Error_Train=mean(abs_error))

preds = predict(model, newdata=data_val)

data_val %>% select(Delivery_Time) %>%
               rename(Actual_Time=Delivery_Time) %>%
               mutate(Predicted_Time = preds) %>%
               mutate(abs_error=abs(Actual_Time-Predicted_Time)) %>%
               summarize(Mean_Absolute_Error_Val=mean(abs_error))
```

Next, scatterplots are created of Actual_Delivery_Time vs. Predicted_Delivery Time.  Ideally, the predicted delivery time should be equal to or greater than the actual delivery time to appropriately manage customer expectations.

The first two plots are for uncalibrated predicted delivery times (second of these plots is faceted by Weekend and Time_of_Day).  From these plots, it's seen that there are many predicted delivery times that are more than the actual delivery times.  The last two plots are for calibrated predicted delivery time (second of these plots is faceted by Weekend and Time_of_Day).  To calibrate, a correction factor is applied to predicted delivery times (post-model prediction) to ensure that they mostly exceed the actual delivery times.  
```{r scatterplots of actual delivery times vs. predicated delivery times (uncalibrated & calibrated)}
df_pred <- data_val %>% 
                 mutate(Predicted_Delivery_Time=preds) %>%
                 rename(Actual_Delivery_Time=Delivery_Time) %>% 
                 mutate(Delivery_Under_Prediction_Time =
                    ifelse(Predicted_Delivery_Time > Actual_Delivery_Time,"Yes","No"))
                  
df_pred %>% ggplot(., aes(x=Predicted_Delivery_Time, y=Actual_Delivery_Time, 
                          color=Delivery_Under_Prediction_Time)) + 
                        geom_jitter(alpha=0.7) + 
                        geom_abline(intercept = 0, slope=1, 
                                    linetype="dashed", size=0.5) + 
                        geom_hline(yintercept=60, linetype="dashed", color = "red", size=0.5) +
                        coord_fixed(ratio=1, xlim=c(0, max(df_pred$Actual_Delivery_Time)),
                                    ylim=c(0, max(df_pred$Actual_Delivery_Time))) +
                        theme_bw() + 
                        labs(x="Predicted Delivery Time (Uncalibrated)", y="Actual Delivery Time") +
                        geom_text(aes(x=200, label="60 minutes", y=60), colour="red",vjust = 1.5) + 
                        ggtitle("Scatterplot of Actual Delivery Time vs. Uncalibrated Predicted Delivery Time")

df_pred %>% ggplot(., aes(x=Predicted_Delivery_Time, y=Actual_Delivery_Time, 
                          color=Delivery_Under_Prediction_Time)) + 
                        geom_jitter(alpha=0.7) + 
                        geom_abline(intercept = 0, slope=1, 
                                    linetype="dashed", size=0.5) + 
                        geom_hline(yintercept=60, linetype="dashed", color = "red", size=0.5) +
                        coord_fixed(ratio=1, xlim=c(0, max(df_pred$Actual_Delivery_Time)),
                                    ylim=c(0, max(df_pred$Actual_Delivery_Time))) +
                        theme_bw() + 
                        labs(x="Predicted Delivery Time (Uncalibrated)", y="Actual Delivery Time") +
                        geom_text(aes(x=170, label="60 minutes", y=65), colour="red",vjust = 1.5) +
                        facet_grid(rows = vars(Weekend), cols=vars(Time_of_Day)) + 
                        ggtitle("Scatterplot of Actual Delivery Time vs. Uncalibrated Predicted Delivery Time")

Calibrate_int=10
Calibrate_slope=1.2

df_pred <- df_pred %>%
               mutate(Predicted_Delivery_Time=Predicted_Delivery_Time*Calibrate_slope+Calibrate_int) %>%
               mutate(Delivery_Under_Prediction_Time=
                  ifelse(Predicted_Delivery_Time > Actual_Delivery_Time,"Yes","No"))

df_pred %>% ggplot(., aes(x=Predicted_Delivery_Time, y=Actual_Delivery_Time, 
                          color=Delivery_Under_Prediction_Time)) + 
                        geom_jitter(alpha=0.7) + 
                        geom_abline(intercept = 0, slope=1, 
                                    linetype="dashed", size=0.5) + 
                        geom_hline(yintercept=60, linetype="dashed", color = "red", size=0.5) +
                        coord_fixed(ratio=1, xlim=c(0, max(df_pred$Actual_Delivery_Time)+Calibrate_int),
                                    ylim=c(0, max(df_pred$Actual_Delivery_Time)+Calibrate_int)) +
                        theme_bw() + 
                        labs(x="Predicted Delivery Time (Calibrated)", y="Actual Delivery Time") +
                        geom_text(aes(x=200, label="60 minutes", y=60), colour="red",vjust = 1.5) + 
                        ggtitle("Scatterplot of Actual Delivery Time vs. Calibrated Predicted Delivery Time")

df_pred %>% ggplot(., aes(x=Predicted_Delivery_Time, y=Actual_Delivery_Time, 
                          color=Delivery_Under_Prediction_Time)) + 
                        geom_jitter(alpha=0.7) + 
                        geom_abline(intercept = 0, slope=1, 
                                    linetype="dashed", size=0.5) + 
                        geom_hline(yintercept=60, linetype="dashed", color = "red", size=0.5) +
                        coord_fixed(ratio=1, xlim=c(0, max(df_pred$Actual_Delivery_Time)+Calibrate_int),
                                    ylim=c(0, max(df_pred$Actual_Delivery_Time)+Calibrate_int)) +
                        theme_bw() + 
                        labs(x="Predicted Delivery Time (Calibrated)", y="Actual Delivery Time") +
                        geom_text(aes(x=180, label="60 minutes", y=65), colour="red",vjust = 1.5) +
                        facet_grid(rows = vars(Weekend), cols=vars(Time_of_Day)) + 
                        ggtitle("Scatterplot of Actual Delivery Time vs. Calibrated Predicted Delivery Time") 
                        
```

## 5) Conclusions

Utilizing the model to predict delivery times and then applying the correction factor allows the vast majority of predicted delivery times to exceed the actual delivery times.  This is a great thing, as customers will likely be more satisfied if they get deliveries sooner than expected.  Though this is not a permanent solution to the problem, it is a great temporary solution while the company sorts out other mechanisms to decrease voids and reduce delivery times.  
