---
title: "Child Mortality and GDP--Merging and Cleaning Prior to Visualization"
author: "Author: Anil Kumar"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1) Introduction

For this project, I needed to clean and merge two datasets that I obtained from the gapminder website: 
https://www.gapminder.org/data/

**Goal:** Clean and merge the two datasets such that each observation (row) represents: 

* Country
* Year
* GDP per capita of the country for the year
* Child mortality rate of the country for the year

The final cleaned dataset will be written to a .csv file and visualization will be completed in Tableau.  

### 2) Merging and Cleaning

```{r read in the necessary libraries, message=FALSE}
# Include the necessary libraries 
# (not all may be used...but I usually include the most common ones I use by habit)
library(tidyverse)
library(readr)
library(readxl)
library(jtools)
library(plotly)
library(caret)
library(pROC)
library(MLmetrics)
library(janitor)
library(lubridate)
```


Both .csv files are read in and named appropriately with "raw" as the suffix.   
```{r read in csv files, message=FALSE}
#read in the data and clean the column names
df_mortality_raw <- read_csv("child_mortality_0_5_year_olds_dying_per_1000_born.csv")
df_gdp_raw <- read_csv("gdppercapita_us_inflation_adjusted.csv")
```

```{r}
head(df_mortality_raw)
```

```{r}
head(df_gdp_raw)
```

It's clear that neither of the datasets are in tidy format.  That is, the years should not be columns.

First, the dataframe containing gdp per capita values is converted to tidy format and saved into a new dataframe.  The raw data is deleted.  
```{r}
df_gdp <- df_gdp_raw %>% pivot_longer(., cols=2:61, names_to="Year", 
                                      values_to="GDP")
rm(df_gdp_raw)
```


Next, the same is done for the dataframe containing child mortality rates.  
```{r}
df_mortality <- df_mortality_raw %>% pivot_longer(., cols=2:302, names_to="Year", 
                                                  values_to="Mortality")

rm(df_mortality_raw)
```


Check to see that both dataframes are now in tidy format.
```{r}
head(df_mortality)
head(df_gdp)
```


Next:

1. The dataframes are merged with an inner join by **country** and **year**.
2. Rows with missing values are dropped
3. **country** column is renamed to **Country**
4. **Year** column is converted to numeric value (previously character)
5. When examning the GDP column, some values are written with a "k" suffix.  For example, **12k** is written instead of **12,000**.  
  + For any cells where this is the case, the "k" is stripped and the value of the cell is multiplied by 1,000 to reflect the actual numeric value
  + All values in the **GDP** column are also converted to numeric (previously character)
6. Data is filtered for years 2001 and later
```{r, message=FALSE, warning=FALSE}
df_gdp_mortality <-df_gdp %>% 
                    inner_join(df_mortality, by=c("country"="country", "Year"="Year")) %>% 
                    drop_na() %>%
                    rename(Country=country) %>%
                    mutate(Year=as.integer(Year)) %>%
                    mutate(GDP = ifelse(
                                substring(GDP,nchar(GDP),nchar(GDP))=="k",
                                as.numeric(substring(GDP,nchar(1),nchar(GDP)-1)) * 1000, 
                                as.numeric(GDP))) %>%
                    filter(Year>=2001)
```

Confirm no missing values and check to see that the final dataframe is in the correct format.  
```{r}
sum(is.na(df_gdp_mortality))
head(df_gdp_mortality)
```

Merged and cleaned dataframe is written to a .csv file.  
Data will be visualized in Tableau!
```{r}
write.csv(df_gdp_mortality,"gdp_and_childmortality_2001_2018.csv", row.names = FALSE)
```


















